---
title: "Using IUCN Spatial Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using IUCN Spatial Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, knitr}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Install and load packages

```{r, library}
# pak::pkg_install("heavywatal/habistats")
library(habistats)
```

## Set options

Directory containing [IUCN shape files](https://www.iucnredlist.org/resources/spatial-data-download):

```{r, options-source}
#| eval: false
# Set
options(habistats.iucn_source = "path/to/your/data")

# Check
habistats::iucn_source_path()
```

You can leave this unset if you want to use the example data included in the package,
which was randomly generated to be similar to the IUCN data.

Intermediate data files will be stored in the cache directory:

```{r, options-cache}
options(habistats.cache_dir = "~/.cache/habistats-example")
```


## Read IUCN shape files

Non-geometry columns are extracted as a pure `data.frame`.
It may take a while to run them for the first time,
but subsequent runs will be very quick because the results are cached.

```{r, features}
features = habistats::iucn_spatial_features() |> print()

species = habistats::iucn_spatial_species() |> print()
```

Use them to filter the species you want to evaluate.

```{r, select-species}
sn = species |>
  dplyr::filter(phylum == "CHORDATA") |>
  dplyr::pull(sci_name) |>
  print()
```


## Preprocess shape files

Split IUCN shape files into separate gpkg files for efficient processing.
On a slow storage, it may take a while to run this even for the non-first time.

```{r, split}
species_gpkg = habistats::iucn_species_gpkg() |> print()
```

## Evaluate species ranges

```{r, evaluate}
out = species_gpkg |>
  dplyr::filter(sci_name %in% sn) |>
  dplyr::pull("source", "sci_name") |>
  quantify_vector_kgc() |>
  print()
```

You may want to save the result to a file.

```{r, write}
#| eval: false
readr::write_tsv(out, "iucn_range.tsv.gz")
```
